Dynamic modulation of social gaze by sex and familiarity in marmoset dyads

Version 1

Abstract

Summary

Social communication relies on the ability to perceive and interpret the direction of others’ attention, and is commonly conveyed through head orientation and gaze direction in humans and nonhuman primates. However, traditional social gaze experiments in nonhuman primates require restraining head movements, significantly limiting their natural behavioral repertoire. Here, we developed a novel framework for accurately tracking facial features and three-dimensional head gaze orientations of multiple freely moving common marmosets (Callithrix jacchus). By combining deep learning-based computer vision tools with triangulation algorithms, we were able to track the facial features of marmoset dyads within an arena. This method effectively generates dynamic 3D geometrical facial frames while overcoming common challenges like occlusion. To detect the head gaze direction, we constructed a virtual cone, oriented perpendicular to the facial frame. Using this pipeline, we quantified different types of interactive social gaze events, including partner-directed gaze and joint gaze to a shared spatial location. We observed clear effects of sex and familiarity on both interpersonal distance and gaze dynamics in marmoset dyads. Unfamiliar pairs exhibited more stereotyped patterns of arena occupancy, more sustained levels of social gaze across social distance, and increased social gaze monitoring. On the other hand, familiar pairs exhibited higher levels of joint gazes. Moreover, males displayed significantly elevated levels of gazes toward females’ faces and the surrounding regions, irrespective of familiarity. Our study reveals the importance of two key social factors in driving the gaze behaviors of a prosocial primate species and lays the groundwork for a rigorous quantification of primate behaviors in naturalistic settings.

Introduction

Primates, including humans, exhibit complex social structures and engage in rich interactions with members of their species, which are crucial for their survival and development. Among social stimuli, the face holds paramount importance with specialized neural systems (Deen et al., 2023; Hesse & Tsao, 2020) and is attentively prioritized by primates during much of their social interaction. Notably, the eyes garner the most attention among all facial features, playing a pivotal role in indicating the direction of others’ attention and also possibly their intention (Dal Monte et al., 2015; Emery, 2000; Itier et al., 2007). Indeed, understanding and interpreting the gaze of fellow individuals is a fundamental attribute of the theory of mind (ToM) (Martin & Santos, 2016; Saxe & Kanwisher, 2003). While current studies of social gaze using either realistic stimuli or pairs of rhesus macaques (Macaca mulatta) in a controlled laboratory setting (Dal Monte et al., 2016; Mosher et al., 2014; Ramezanpour & Thier 2020; Shepherd et al., 2006; Shepherd & Freiwald, 2018) provide valuable insights into social gaze behaviors, they are nevertheless limited in their ecological relevance.

To address these limitations, we turned to common marmosets (Callithrix jacchus), a highly prosocial primate species known for their social behavioral and cognitive similarities to humans (Miller et al., 2016). Marmosets are also a model system with increasing applications in computational ethology (Mitchell et al., 2014; Ngo et al., 2022). Like humans, they engage in cooperative breeding, a social system in which individuals care for offspring other than their own, usually at the expense of their reproduction (French, 1997; Solomon & French, 1997). Gaze directions, inferred from head orientation, hold crucial information about marmoset social interactions (Heiney & Blazquez, 2011; Spadacenta et al., 2022). The emergence of computational ethology (Anderson & Perona, 2014; Datta, Anderson, Branson, Perona, & Leifer, 2019) has propelled the development of a host of computer vision tools using deep neural networks (e.g., OpenPose by Cao et al., 2017, DeepLabCut by Mathis et al., 2018, DANNCE by Dunn et al., 2021; SLEAP by Pereira et al., 2022). Moreover, we have recently developed a scalable marmoset apparatus for automated pulling (MarmoAAP) for studying cooperation in marmoset dyads (Meisner et al., 2024). However, tracking head gaze direction in multiple freely moving marmosets poses a challenging problem, not yet solved by existing computer vision tools. This problem is further complicated by the fact that accurately tracking gaze orientations in primates requires three-dimensional information.

Here, we propose a novel framework based on a modified DeepLabCut pipeline, capable of accurately detecting body parts of multiple marmosets in 2D space and triangulating them in 3D space. By reconstructing the face frame with six facial points in 3D space, we can infer the head gaze of each animal across time. Importantly, marmosets that are not head-restrained use rapid head movements for reorienting and visual exploration (Pandey et al., 2020), and therefore the head direction serves as an excellent proxy for gaze orientation in unrestrained marmosets. With this framework in place, we investigated the gaze behaviors of male-female pairs of freely moving and interacting marmosets to quantify their social gaze dynamics. We investigated these gaze dynamics along the dimensions of sex and familiarity and found several key differences along both of these important social dimensions, including increased partner gaze in males that is modulated by familiarity, increased joint gaze among familiar pairs, and increased gaze monitoring by males. This fully automated tracking system can thus serve as a powerful tool for investigating primate group dynamics in naturalistic environments.

Results

Experimental setup and reconstruction of video images in 3D space

The experimental setup consisted of two arenas made with acrylic plates that allowed two marmosets to visually interact with each other while being physically separate (Fig. S1A). Each arena was 60.96 cm long, 30.48 cm wide, and 30.48 cm high. Five sides of the arena, except the bottom, were transparent allowing a clear view of the animal subjects under observation. The bottom side of the arena was perforated with 1-inch diameter holes arranged in a hexagonal pattern to aid the animal’s traction. The arenas were mounted on a rigid frame made of aluminum building blocks, with the smaller sides facing each other, and were separated by a distance of 30.48 cm. A movable opaque divider was placed between the arenas during intermittent breaks to prevent the animals from having visual access to each other (Methods). Two monitors were attached to the aluminum frame, one on each end, for displaying video or image stimuli to the animals. To capture the whole experimental setup, two sets of four GoPro 8 cameras were attached to the frame, where each set of cameras captured the view of one of the arenas.

After obtaining the intrinsic parameters of the cameras by calibration and the extrinsic parameters of the cameras by L-frame analysis (see Methods), we established a world coordinate system of each arena surrounded by the corresponding set of four cameras. Crucially, the two independent world coordinate systems of the two arenas were combined by measuring the distance between the two L-shaped frames and adding this offset to one of the world coordinate systems.

With the established world coordinate system, any point captured by two or more cameras could be triangulated into a common three-dimensional space. Thus, the experimental setup was reconstructed into three-dimensional space by manually labeling the vertices of the arenas and monitors in the image space captured by the cameras (Fig. S1B). The cameras on the monitor ends (marked as ‘ML1’, ‘ML2’, ‘MR1’, ‘MR2’ in Fig. S1B) recorded both animal subjects, whereas the cameras in the middle (marked as ‘OL1’, ‘OL2’, ‘OR1’, ‘OR2’ in Fig. S1B) recorded only one animal subject.

Automatic detection of facial features of two marmosets

Six facial features – the two tufts, the central blaze, two eyes, and the mouth (Fig. 1A) – were selected for automated tracking using a modified version of a deep neural network (DeepLabCut, Mathis et al., 2018). The raw video from each camera was fed into the network to compute the probability heatmaps of each facial feature. We modified the original method to detect features of two animals (Fig. 1B). After processing the raw video, two locations with the highest probability over a threshold (95%) were picked from each probability heatmap (Fig. 1B, feature detection). Since all the features from the same animal should be clustered in image space, a K-means clustering algorithm (Fig. 1B, initial clustering) was used on the candidate features with the constraint that one animal can only have one unique feature (Fig. 1B, refine clustering); for example, one animal cannot have two left eyes. After clustering, two clusters of features corresponding to the two animals were obtained. To detect outliers that were not valid features, we first calculated a distribution of within-cluster distances (Fig. 1B, remove outliers). Outliers were determined as those points that had nearest-neighbor distances which were two standard deviations above the average within-cluster distance and were excluded from subsequent analyses. Note that the above analyses were performed independently for each video frame.

Pipeline of detecting facial features of two marmosets.

(A) Six facial features of the marmoset (face frame) are color-coded: right tuft (red), central blaze (yellow), left tuft (green), right eye (purple), left eye (blue), and mouth (magenta). (B) Feature tracking pipeline (right) with the corresponding illustration for each step across two adjacent video frames (left). At the end of the pipeline, the facial features are clustered with the identities assigned consistently across frames. Facial points are color-coded as in A. (C) Example frames of four steps in the pipeline shown in B. It can be seen clearly that the facial points are tracked and clustered accurately, and the identities are consistent across frames.

To establish temporal continuity across video frames and track animal identities, we first calculated the centroid of each cluster (Fig. 1B, calculate centroids). Under the heuristic that centroid trajectories corresponding to each individual animal are smoothly continuous in space and time (i.e., there are no sudden jumps or reversals in centroid location across frames at our sampling rate of 30Hz), we assigned identities to the centroids thus enabling us to track identities over time (Fig. 1B, establish identities). The facial features corresponding to a centroid inherited this identity (Fig. 1B, cluster with identities). We tested our automated pipeline against a manually labeled data set and found that features were detected with high precision (root mean squared error = 3.31 ± 0.1 pixels [mean +/-s.e.m.]; n = 2400 facial features; see Methods).

Our method thus allowed us to accurately detect and track the facial features of two marmosets in the arena (Fig. 1C), and in more general contexts such as in a home cage with occlusions (Supplementary Video 1).

Inferring head-gaze direction in 3D space

In our experimental setup, the cameras in the middle unambiguously recorded only one animal. The centroids of the facial features in the image space recorded by the middle cameras were triangulated into three-dimensional space and were constrained to be confined within the bounds of the arena. Any missing centroids were filled by interpolation from neighboring frames from both the past and future time points. The triangulated centroids acquired from cameras in the middle were then projected into the image space of cameras on the monitor ends. Since both animals were recorded by the cameras on the monitor ends, facial features from these camera views were detected with identities assigned (as described in the 2D pipeline). In the image space of cameras on the monitor ends, the cluster of facial feature points closer to the projected centroid and within the bounds of the arena were kept for later triangulation.

The detected facial features captured by all four cameras for one animal were subjected to triangulation. For each feature, results from all possible pairs of four cameras were triangulated. All the triangulation results were averaged to yield the final coordinates of the body part in three-dimensional space. Any missing features were filled by interpolation from neighboring frames including the previous and future time points. Testing against a manually labeled and triangulated data set yielded a high precision of 3D feature detection (root mean squared error = 3.7 ± 0.1 mm [mean +/-s.e.m.]; n = 2400 facial features; see Methods).

The six facial points constituted a semi-rigid geometric frame as the animal moves in 3D space (‘face frame’; Fig. 1), allowing us to infer the animal’s head-gaze direction as follows. The gaze orientation was calculated as the normal to the facial plane (‘face norm’) defined by the two eyes and the central blaze. The position of the ear tufts, which were behind this facial plane, was used to determine gaze direction. Since marmoset saccade amplitudes are largely restricted to 10 degrees (median less than 4 degrees) (Mitchell et al., 2014), we modeled the head gaze as a virtual cone with a solid angle of 10 degrees (‘gaze cone’) emanating from the facial plane (Fig. 2A). Notably, with multiple camera views, the face frame can be reconstructed even when the face was invisible to one of the cameras, such that the reconstructed face frame in 3D can be projected back into the image space to validate the accuracy of the detection and reconstruction (Fig. 2B). We were thus able to obtain the animal’s continuous movement trajectory and the corresponding gaze direction over time (Fig. 2C, Supplementary Video 2). To characterize stable gaze epochs for subsequent analyses, we calculated the velocity of the face norm and applied a stringent threshold of 0.05 (normalized units) below which the marmoset’s head movement was considered stationary (Fig 2D).

3D Reconstruction of facial features and head gaze modeling.

(A) The face frames of two marmosets are reconstructed in 3D using the tracked facial points in Fig. 1 A. A cone perpendicular to the face frame (gaze cone; 10-degree solid angle) is modeled as the head gaze. (B) Two example frames with the facial points projected from 3D space onto different camera views are shown. The left frame demonstrates that the facial points can be detected using information from other cameras, even if the face is invisible from that viewpoint. (C) Trajectory of the reconstructed face frame and the corresponding gaze cones across time. (D) The histogram of the face norm velocity. The red dotted line (0.05) is the threshold below which the marmoset head direction is considered to be stationary.

We first examined our method’s ability to characterize gaze behaviors of freely moving marmosets by presenting either video or image stimuli to individual animals on a monitor screen (see Methods). Marmosets exhibited longer gaze duration to video stimuli compared to image stimuli (Fig. S2A; Mann Whitney U Test, p < 0.001). However, this difference was not caused by differences in gaze dispersion (Fig. S2B; Mann Whitney U Test, ns). By examining the frequency of gaze events, we found that marmosets gazed at the monitor more during the early period of video stimuli presentations compared to the late period (Fig. S2C; Mann Whitney U Test, p < 0.001), while there was no such difference for the image stimuli (Fig. S2C; Mann Whitney U Test, p = 0.4345). There was also a significant difference in gaze frequency between the early period of video presentations compared to the same period for image presentations (Fig. S2C; Mann Whitney U Test, p < 10−10). Taken together, our results support that dynamic visual stimuli elicit greater overt attention compared to static stimuli in marmosets, similar to macaques (Dal Monte et al., 2016; Furl et al., 2012) and humans (Chevallier et al., 2015).

Positional dynamics of marmoset dyads

With this automated system in place, we recorded the behavior of four pairs of familiar marmosets and four pairs of unfamiliar marmosets. Data from each pair was recorded in one session consisting of ten five-minute free-viewing blocks interleaved with five-minute breaks. Each pair consisted of a male and a female animal. The familiar pairs were defined as cage mates, while each member of an unfamiliar pair was from a different home cage with no visual access to each other while in the colony. We first examined the movement trajectories of marmoset dyads and used the centroids of the face frames across time to represent the trajectories. For an example 5-minute segment (Fig. 3A), we observed that the marmosets preferred to stay at the two ends of their respective arenas. This was confirmed by a heatmap of projections of the trajectories on the plane parallel to the vertical long side of the arenas (‘XZ’ plane). Furthermore, there were two hotspots along the vertical axis in the heatmap of projections to the vertical short side plane (‘YZ’ plane), suggesting that the animals’ preferred body postures were either upright or crouched.

Positional dynamics of marmoset dyads

(A) Movement trajectories of the face frame centroids for a marmoset pair (orange for female, cyan for male) in an example five-minute block. The heatmaps were calculated using the projections of the trajectories to XY, YZ, and XZ planes. (B) Marginal distributions of movement trajectories along the X and Z axes were calculated for all marmosets and grouped by familiarity and sex (transparent colors for familiar pairs, opaque colors for unfamiliar pairs). Black bars indicate significant differences between pairs of distributions (Mann-Whitney U test, significance level at 5%). (C) Histograms of social distance along the X axis show trimodal distributions for both familiar pairs (gray) and unfamiliar pairs (black). The top inset shows a schematic of the arena configuration with corresponding colors (orange for females and cyan for males). Social distance is defined as the Euclidean distance between the two marmosets. I he fitted red curve for the unfamiliar pairs is a tn-Gaussian distribution, while the fitted red curve for the familiar pairs is a mixture of Gamma and Gaussian distributions, with the first peak as the Gamma distribution. The three regions were designated as ‘Near’, Intermediate’, and ‘Far’. The bottom inset illustrates the reason for this nomenclature. (D) Temporal evolution of social distance for unfamiliar and familiar pairs (which each 5-minute viewing block). The central dark line is the mean and the shaded area is the standard deviation. Black dots indicate significant differences (Mann-Whitney U test, significance level at 5%).

To quantify their positional dynamics, we examined the marginal distributions of the movement trajectories along the horizontal (‘X’) and vertical (‘Z’) axes across all sessions and grouped them along the dimensions of sex and familiarity (Fig. 3B). Along the X axis (Fig. 3B, left), the distributions were slightly bimodal, with the main peak in the region near to the inner edge of the arenas. Regardless of familiarity, male marmosets tended to stay closer to the inner edge compared to females, as shown by the significant differences in the distributions when X ranged from 0 to 150 mm. However, there were no significant differences between the same sex members of familiar and unfamiliar pairs. For the Z axis (Fig. 3B, right), the distributions were bimodal (Bimodality coefficients (BC) all exceeded the threshold at 5/9 with a 5% margin; BC(familiar male) = 0.5863; BC(familiar female) = 0.6278; BC(unfamiliar male) = 0.6235; BC(unfamiliar female) = 0.6570; Warren Sarle’s bimodality test), consistent with what we observed in the heatmaps indicating either upright or crouched postures. The positional distributions along the Z axis were not different based on sex or familiarity.

To further characterize the “social distance” dynamics of the freely moving dyads, we calculated the distance between the centroids of the pairs. We then examined the distributions of the social distance along the X-axis separately for familiar and unfamiliar pairs (Fig. 3C). The distributions were trimodal, and can be explained by the bimodal distribution of movement trajectories of individual marmosets along the X-axis. As mentioned above, marmosets tended to stay at the two ends of their arenas, and thus, combinations of preferred positions at the two ends for the dyads (see insets in Fig. 3C) resulted in the trimodal distribution. We termed these three peaks as ‘Near’, ‘Intermediate’, and ‘Far’. To quantify these distributions, we fitted the empirical data with mixture models using maximum likelihood estimation (see Methods). The social distance for unfamiliar pairs was best fitted by a mixture of three Gaussians while the distribution for familiar pairs was best fitted by a mixture of Gamma and Gaussian distributions. The first peak (‘Near’) of the familiar-pair distribution was best fitted by a Gamma distribution, implying a higher degree of dispersion when familiar marmosets were close to each other. Upon examining the temporal evolution of the social distance (within each 5-minute viewing block), we found that the social distance of unfamiliar pairs increased over time, whereas this distance fluctuated over time for familiar pairs (Fig. 3D), further indicating that the social distance dynamics of marmoset dyads depended on familiarity.

Social gaze dynamics of marmoset dyads

We next investigated the interactive aspects of gaze behaviors in freely moving marmoset dyads. The gaze interaction between two animals could be simplified as the relative positions of two gaze cones in three-dimensional space (see Methods; Fig. 4A; Supplementary Video 3). If the gaze cone of one animal intersected with the facial plane of the second animal (but not vice versa), we termed it ‘partner gaze’. If the gaze cones of both animals intersected with that of the other’s facial place, we termed it ‘reciprocal gaze’. In our dataset, the instances of reciprocal gaze were very low and were thus excluded from further analysis. If the two cones intersected anywhere outside the facial planes, we termed it ‘joint gaze’. All other cases were regarded as ‘no interaction’ between the two animals.

Live interactive gaze analysis of unfamiliar and familiar marmoset dyads

(A) Gaze type categorized based on the relative positions of the gaze cones Joint gaze is defined as two marmosets looking at the same location. A partner gaze is defined as one animal looking at the other animal’s face (but not vice versa) No interaction occurs when the two gaze cones do not intersect (B) Histograms of gaze count as a function of inter-animal distance, shown separately for familiar and unfamiliar pairs (C) Same data as m B shown as pie charts of percentages in social gaze states (D) Gaze state transition diagrams for familiar and unfamiliar pairs The nodes ate the gaze states and the edges connecting the nodes represent the transition between states Edge colors indicate transition probabilities (E) Partner gaze self-transition probabilities for familiar and unfamiliar pairs (χ2 test) (F) Delta transition matrix between the unfamiliar pair and familiar pair state transition diagrams. Transitions that are significantly different across familiarity are marked by astensks (χ2 test, male to male, p < 0.05; female to male, p < 0.05; joint to joint p < 0.0001). (G) Left. The schematic illustrates hew gazing toward the surrounding region of a partner’s face area was measured Right Counts of gaze towards the surrounding region of the partner s face by familiarity and sex. (Mann-Whitney U test, **** means p < 0.001. **** means p < 0.0001)

We identified stable gaze epochs by thresholding the face norm velocity (Fig. 2D). Stable epochs were categorized into gaze states based on the gaze event types described above. We first analyzed the fraction of gaze states in the three position ranges identified from the social distance analysis (Fig. 4B,C). We found that male marmosets gazed more toward their partner females’ faces regardless of familiarity (p < 0.01, χ2 test). Fraction of male → female partner gaze incidents decreased with increasing social distance for familiar pairs, while they remained constant in the case of unfamiliar pairs (Fig. 4C). Moreover, females in unfamiliar pairs exhibited significantly higher partner gazes (female→male) compared to those in familiar pairs (p < 0.01, χ2 test; Fig. 4C). The total counts of social gaze states (joint gaze and partner gaze) were higher for familiar pairs when they were near, but these decreased more dramatically with increasing distance (Fig. 4B).

To investigate the dynamics of these gaze states, we computed state transition probabilities among distinct gaze event types for familiar and unfamiliar dyads. We applied a Markov chain model (see Methods) (Fig. 4D), in which the nodes were the gaze states and the edges connecting the nodes represented the transitions between gaze states. We first focused on the recurrent (self-transition) edges for the partner gaze states. Recurrent edges indicate a transition back to the same stable gaze state after a break likely due to physical movement, and reflect the robustness of the state despite movement. In line with our previous results (Fig. 4B,C), males exhibit significantly higher recurrent partner gazes compared to females, irrespective of familiarity (χ2 test, unfamiliar male vs unfamiliar female, p < 10−10; familiar male vs familiar female, p < 0.01; Fig. 4E).

A comparison of state transition probabilities across the dimension of familiarity yielded several noteworthy findings (Fig. 4F). First, recurrent male partner gaze (male female) was significantly enhanced in unfamiliar pairs (p < 0.05, χ2 test), suggesting a heightened interest in unfamiliar females. Second, there was a higher probability of transition from a female partner gaze to a male partner gaze in familiar pairs compared to unfamiliar pairs, suggesting that familiar males have a greater awareness of and tendency to reciprocate their partners’ gaze (p < 0.05, χ2 test). Third, there was a higher probability of recurrent joint gazes in familiar pairs compared to unfamiliar pairs, suggesting that familiar pairs explore common objects more than unfamiliar pairs (p < 10−4, χ2 test).

Monitoring others to anticipate their future actions is critical for successful social interactions (Hari et al., 2015). In particular, successful interactive gaze exchanges require constant monitoring of other’s gaze. We analyzed the gaze distribution in the surrounding region of a partner’s face to estimate gaze monitoring tied to increased social attention (Dal Monte et al., 2022). We quantified this by the distance between the centroid of the partner’s face-frame and the point of intersection of the gaze cone with the partner’s facial plane (Fig. 4G, left). Unfamiliar marmosets (both males and females) showed significantly higher (Mann-Whitney U test, unfamiliar male vs familiar male, p < 0.0001; unfamiliar female vs familiar female, p < 0.001) incidences of gaze toward the surrounding region of the partner’s face (Fig 4G, right; compare darker lines with the lighter lines). Further, males exhibited markedly higher (Mann-Whitney U test, p < 0.0001) incidences of gaze toward the partner females’ face (Fig 4G, right; compare cyan lines with the orange lines).

Joint gaze is crucial in primates as it underpins social communication and coordination, serving as a foundation for more complex behaviors like cooperation and shared attention (Emery, 2000; Tomasello et al., 2005). We next analyzed joint gaze behaviors between marmoset dyads, first projecting the locations of joint gazes onto different 2D planes around the arena (Fig. 5A). Significant asymmetry was observed in the distribution of joint gazes on the ‘XY’ and ‘XZ’ planes, with the majority of joint gazes occurring within the female’s arena, regardless of familiarity. This is consistent with our earlier results showing more partner-directed gazes from males and increased attention to the region surrounding the female partner (Fig. 4B,C,G). Notably, joint gazes between familiar pairs were more centrally distributed compared to those between unfamiliar pairs. When examining the ‘XZ’ and ‘YZ’ planes, we found that most joint gazes were concentrated along the lower Z-axis, indicating that marmoset pairs tend to establish shared attention closer to ground level. Given the importance of social distance in shaping marmoset behaviors, we further investigated the distribution of joint gazes as a function of social distance. Striking differences were found between familiar and unfamiliar pairs (Fig. 5B). The distribution for familiar pairs was best fitted by a tri-Gaussian model, consistent with the distribution of social distances observed in Fig. 3C, suggesting that familiar pairs can establish joint gazes regardless of distance. In contrast, the distribution for unfamiliar pairs was best described by a lognormal model, indicating that unfamiliar pairs tend to establish joint gazes only when in close proximity.

Joint gaze analysis between familiar and unfamiliar dyads.

(A) Heatmaps of joint gaze locations projected onto 2D planes (From left to right: ‘XY’, ‘XZ’, ‘YZ’). The icon on the left show a schematic of a joint gaze along with the arena configuration with corresponding colors (orange for females and cyan for males). The colored rectangles superimposed on the heatmaps show the projections of the arenas onto those 2D planes. The white dotted line indicates the midline between the two arenas. (B) The probability of joint gaze at varying social distances shows distinct distributions for unfamiliar pairs (black) and familiar pairs (gray). The fitted red curve for the unfamiliar pairs is a lognormal distribution, while the fitted red curve for the familiar pairs is a tri-Gaussian distribution. Thin fitted red curves in both panels are the data from individual pairs. The inset shows a schematic of a joint gaze and the social distance metric. Social distance is defined as the Euclidean distance between the two marmosets

Overall, we found that both the social dimensions we examined – familiarity and sex – are significant determinants of natural gaze dynamics among marmosets.

Discussion

In this study, we first presented a novel framework for the automated, markerless, and identity-preserving tracking of 3D facial features of multiple marmosets. By building on top of the deep-learning framework provided by DeepLabCut, we used a constellation of cameras to overcome “blindspots” due to occlusion and imposed spatiotemporal smoothness constraints on the detected features to establish and preserve identities across time. The tracked facial features from each animal form a semi-rigid face frame as the animal moves freely in 3D space, thereby allowing us to infer the animal’s gaze direction at each moment in time. It is important to reiterate that unrestrained marmosets use rapid saccadic head-movements for reorienting (Pandey et al., 2020) and have a limited amplitude range of saccadic eye-movements (Mitchell et al., 2014). Thus their head direction serves as excellent proxy for gaze orientation in unrestrained conditions.

Current methods for tracking and estimating animal body postures, such as DeepLabCut (Mathis et al., 2018) and SLEAP (Pereira et al., 2022), offer substantial advantages in precision and flexibility, particularly through the use of deep learning for markerless tracking across diverse species. These methods excel in laboratory settings, where they can achieve high accuracy without invasive markers and are adaptable for various animals and tasks. However, many of these tools struggle with occlusion, especially in crowded environments or natural habitats, limiting their effectiveness in complex social interactions or open-field studies (Sturman et al., 2020). Additionally, when tracking marmosets that exhibit significant vertical movement, requiring 3D tracking, the performance of these methods is suboptimal. While 3D tracking systems like DANNCE (Dunn et al., 2021) offer enhanced spatial accuracy, these methods are limited to single-animal tracking, which is insufficient for studies in social neuroscience.

Primates are highly visual species whose physical explorations of their environment are not confined to two-dimensional surfaces. Gaze is a critical component of primate social behavior and conveys important social signals such as interest, attention, and emotion (Emery, 2000). Assessment of gaze is therefore important to understand non-verbal communication and interpersonal dynamics. Our 3D gaze tracking approach was able to capture both the positional and gaze dynamics of freely moving marmoset dyads in a naturalistic context. We observed clear effects of sex and familiarity on both interpersonal and gaze dynamics. Unfamiliar pairs exhibited more stereotyped patterns of arena occupancy, more sustained levels of social gaze across distance, and increased gaze monitoring, suggesting elevated levels of social attention compared to familiar pairs. On the other hand, familiar pairs exhibited more recurrent joint gazes in the shared environment compared to unfamiliar pairs. Familiar males also showed a higher tendency to reciprocate their partner’s gaze, suggesting a greater awareness of their partner’s social gaze state.

Supported by the natural ethology of marmosets (Yamamoto et al., 2014; Solomon & French, 1997), we found dramatic sex differences in gaze behaviors, with males exhibiting significantly elevated levels of gaze toward females’ faces and the surrounding regions irrespective of familiarity. It is important to note that dominance in marmosets is not strictly determined by gender, as it can vary based on individual personalities and intra-group social dynamics, although breeding females typically dominate social activity within a group (Digby, 1995; Mustoe et al., 2023). While we have not explicitly controlled for dominance in this study, whether part of the observed differences can be attributed to dominance effects needs further exploration.

Gaze following plays a crucial role in social communication for humans and nonhuman primates, allowing for joint attention (Emery et al., 1997; Brooks & Meltzoff, 2005; Burkart & Heschl, 2006; Shepherd, 2010). Previous research demonstrated that head-restrained marmosets exhibited preferential gazing toward marmoset face stimuli observed by a conspecific in a quasi-reflexive manner during a free-choice task (Spadacenta et al., 2019). Interestingly we did not find any differences in gaze-following behaviors (transition from partner gaze to joint gaze) along the social dimensions we tested here. Future investigation of such behaviors by manipulating social variables such as dominance or kinship could provide a comprehensive understanding of gaze following and joint attention in naturalistic behavioral contexts. The scarcity of reciprocal gazes in our study may be attributed to the task-free experimental setup employed. Indeed, in other joint action tasks requiring cooperation for rewards, marmosets actively engage in reciprocal gaze behaviors (Miss & Burkart, 2018).

While we focused on the tracking of facial features in this study, our automated system has the potential to extend to 3D whole-body tracking, encompassing limbs, tail, and the main body features of marmosets. In our system, multiple cameras surrounding the arena ensure that each body part of interest can be tracked through at least two cameras, enabling triangulation in 3D space. Our current system uses a pre-trained ResNet model (He et al., 2015) to track body parts of interest. However, considering the challenges posed by whole-body tracking, such as interference from marmosets’ fur that complicates feature detection, the adoption of cutting-edge transformer networks like the vision transformer model (Dosovitskiy et al., 2020) might significantly improve detection performance. Such an advancement in tracking and reconstructing the entire marmoset body frame would enable the analysis of such data using unsupervised learning techniques (Berman et al., 2014; Calhoun et al., 2019) and thereby provide a deeper understanding of primate social behavior.

In summary, our study lays the groundwork for a rigorous quantification of primate behaviors in naturalistic settings. Not only does this allow us to gain deeper insights beyond what is possible from field notes and observational studies, but it is also a key first step to go beyond current reductionist paradigms and understanding the neural dynamics underlying natural behaviors (Miller et al., 2022).

eLife Assessment

This study establishes the methodology (machine vision and gaze pose estimation) and behavioral apparatus for examining social interactions between pairs of marmoset monkeys. Their results enable unrestrained social interactions under more rigorous conditions with detailed quantification of position and gaze. It has been difficult to study social interactions using artificial stimuli, as opposed to genuine interactions between unrestrained animals. This study makes an important contribution to studying social neuroscience within a laboratory setting; the approach is novel and well-executed, backed by convincing evidence.

Reviewer #1 (Public review):

Summary:

The current study by Xing et al. establishes the methodology (machine vision and gaze pose estimation) and behavioral apparatus for examining social interactions between pairs of marmoset monkeys. Their results enable unrestrained social interactions under more rigorous conditions with detailed quantification of position and gaze. It has been difficult to study social interactions using artificial stimuli, as opposed to genuine interactions between unrestrained animals. This study makes an important contribution for studying social neuroscience within a laboratory setting that will be valuable to the field.

Strengths:

Marmosets are an ideal species for studying primate social interactions due to their prosocial behavior and the ease of group housing within laboratory environments. They also predominantly orient their gaze through head movements during social monitoring. Recent advances in machine vision pose estimation set the stage for estimating 3D gaze position in marmosets but require additional innovation beyond DeepLabCut or equivalent methods. A six-point facial frame is designed to accurately fit marmoset head gaze. A key assumption in the study is that head gaze is a reliable indicator of the marmoset's gaze direction, which will also depend on the eye position. Overall, this assumption has been well supported by recent studies in head-free marmosets. Thus the current work introduces an important methodology for leveraging machine vision to track head gaze and demonstrates its utility for use with interacting marmoset dyads as a first step in that study.

Weaknesses:

One weakness that should be easily addressed is that no data is provided to directly assess how accurate the estimated head gaze is based on calibrations of the animals, for example, when they are looking at discrete locations like faces or video on a monitor. This would be useful to get an upper bound on how accurate the 3D gaze vector is estimated to be, for planned use in other studies. Although the accuracy appears sufficient for the current results, it would be difficult to know if it could be applied in other contexts where more precision might be necessary.

Reviewer #2 (Public review):

Summary:

The manuscript describes novel technique development and experiments to track the social gaze of marmosets. The authors used video tracking of multiple cameras in pairs of marmosets to infer head orientation and gaze and then studied gaze direction as a function of distance between animals, relationships, and social conditions/stimuli.

Strengths:

Overall the work is interesting and well done. It addresses an area of growing interest in animal social behavior, an area that has largely been dominated by research in rodents and other non-primate species. In particular, this work addresses something that is uniquely primate (perhaps not unique, but not studied much in other laboratory model organisms), which is that primates, like humans, look at each other, and this gaze is an important social cue of their interactions. As such, the presented work is an important advance and addition to the literature that will allow more sophisticated quantification of animal behaviors. I am particularly enthusiastic with how the authors approach the cone of uncertainty in gaze, which can be both due to some error in head orientation measurements as well as variable eye position.

Weaknesses:

There are a few technical points in need of clarification, both in terms of the robustness of the gaze estimate, and possible confounds by gaze to non-face targets which may have relevance but are not discussed. These are relatively minor, and more suggestions than anything else.

Author response:

The following is the authors’ response to the original reviews

Public Reviews:

Reviewer #1 (Public review):

Summary:

The current study by Xing et al. establishes the methodology (machine vision and gaze pose estimation) and behavioral apparatus for examining social interactions between pairs of marmoset monkeys. Their results enable unrestrained social interactions under more rigorous conditions with detailed quantification of position and gaze. It has been difficult to study social interactions using artificial stimuli, as opposed to genuine interactions between unrestrained animals. This study makes an important contribution for studying social neuroscience within a laboratory setting that will be valuable to the field.

Strengths:

Marmosets are an ideal species for studying primate social interactions due to their prosocial behavior and the ease of group housing within laboratory environments. They also predominantly orient their gaze through head movements during social monitoring. Recent advances in machine vision pose estimation set the stage for estimating 3D gaze position in marmosets but require additional innovation beyond DeepLabCut or equivalent methods. A six-point facial frame is designed to accurately fit marmoset head gaze. A key assumption in the study is that head gaze is a reliable indicator of the marmoset's gaze direction, which will also depend on the eye position. Overall, this assumption has been well supported by recent studies in head-free marmosets. Thus the current work introduces an important methodology for leveraging machine vision to track head gaze and demonstrates its utility for use with interacting marmoset dyads as a first step in that study.

Weaknesses:

One weakness that should be easily addressed is that no data is provided to directly assess how accurate the estimated head gaze is based on calibrations of the animals, for example, when they are looking at discrete locations like faces or video on a monitor. This would be useful to get an upper bound on how accurate the 3D gaze vector is estimated to be, for planned use in other studies. Although the accuracy appears sufficient for the current results, it would be difficult to know if it could be applied in other contexts where more precision might be necessary.

Please see our detailed responses to the reviewer comments below.

Reviewer #2 (Public review):

Summary:

The manuscript describes novel technique development and experiments to track the social gaze of marmosets. The authors used video tracking of multiple cameras in pairs of marmosets to infer head orientation and gaze and then studied gaze direction as a function of distance between animals, relationships, and social conditions/stimuli.

Strengths:

Overall the work is interesting and well done. It addresses an area of growing interest in animal social behavior, an area that has largely been dominated by research in rodents and other non-primate species. In particular, this work addresses something that is uniquely primate (perhaps not unique, but not studied much in other laboratory model organisms), which is that primates, like humans, look at each other, and this gaze is an important social cue of their interactions. As such, the presented work is an important advance and addition to the literature that will allow more sophisticated quantification of animal behaviors. I am particularly enthusiastic with how the authors approach the cone of uncertainty in gaze, which can be both due to some error in head orientation measurements as well as variable eye position.

Weaknesses:

There are a few technical points in need of clarification, both in terms of the robustness of the gaze estimate, and possible confounds by gaze to non-face targets which may have relevance but are not discussed. These are relatively minor, and more suggestions than anything else.

Please see our detailed responses to the reviewer comments below.

Reviewer #1 (Recommendations for the authors):

Major comments:

(1) It appears that the accuracy of the estimated gaze angle must be well under the size of the gaze cone (+/- 10 degrees), but I can't find any direct estimate of the accuracy even if it is just a ballpark figure. On Lines 219-233 is where performance is described for viewing images and video on a monitor, where it should be possible to reconstruct the point of gaze on the monitor while images and video are shown, in order to evaluate the accuracy of the system for where the marmoset is looking? Would you see eye position traces that would show fixation clusters around those images or videos with stationary points on the monitor much like that seen for head-fixed animals looking at faces on a screen (Mitchell et al, 2014)? If so, what is the typical spread of those clusters during fixations on an image, both in terms of the precision by RMS error during a fixation epoch and the spread around the images at different locations (accuracy of projection)? For example, if gaze clusters were always above the displayed images one would have an idea that the face plane is slightly offset above the true gaze direction. It is not completely clear how well the face plane and corresponding gaze cone do in describing gaze direction in space, but the monitor stimuli could be used as an initial validation of it.

We thank the reviewer for this important suggestion regarding the quantitative validation of gaze accuracy. We agree that, when animals view stimuli presented on a monitor, the estimated gaze direction can be evaluated by examining the spatial distribution of gaze–monitor intersection points relative to stimulus locations.

To address this, we generated a new figure (Fig. S2A) analyzing gaze behavior following the onset of video stimuli presented at different locations on the monitor. Specifically, we selected video clips in which human annotators verified that the marmosets were looking at the monitor. Consistent with prior work in head-fixed marmosets (Mitchell et al., 2014), we observe clustering of gaze–monitor intersection centers within and around the corresponding stimulus locations after stimulus onset. These clusters provide an empirical validation that the estimated gaze direction aligns with stimulus position in space.

Importantly, unlike the head-fixed preparation used in Mitchell et al. (2014), marmosets in our study were freely moving. As a result, they do not exhibit prolonged, stationary fixations on the monitor, and fixation clusters are therefore more diffuse. This increased spread reflects natural head and body motion rather than limitations of the gaze estimation method itself. Despite this, gaze intersection points remain spatially localized to the vicinity of the presented stimuli across different monitor locations.

We did observe small offsets in some gaze clusters relative to stimulus centers; however, these offsets were not systematic across stimulus locations or animals. Crucially, there was no consistent bias (e.g., clusters appearing uniformly above or below stimuli) that would indicate a systematic misalignment of the face plane or gaze cone relative to true gaze direction. Together, these observations support the conclusion that the face-plane-based gaze cone provides an accurate estimate of gaze direction in space, with precision well within the ±10° aperture of the gaze cone.

While the freely moving component of the behavior precludes direct estimation of fixation RMS error comparable to head-fixed paradigms, the observed stimulus-locked clustering serves as an initial validation of both the accuracy and practical utility of our approach under naturalistic conditions.

(2) A second major comment is about clarity in the writing of the results and discussion. At the end of the manuscript, a major takeaway is the difference between familiar and unfamiliar dyads, that males show more interest in viewing females including unfamiliar females, but for familiar females, this distinction is also associated with being likely to look at them if they look at the male, and then to engage in joint gaze with them after looking at them, which indicates more of a social interaction than simply monitoring them when they are unfamiliar. Those aspects of the results could be emphasized more in the topic sentences of paragraphs presenting data to support those features of the gaze data (at present is buried at the ends of results paragraphs and back in the discussion).

We thank the reviewer for this insightful suggestion. We have restructured the Results and Discussion sections to lead with the primary social takeaways rather than technical descriptions (Tracked changes in Word). Specifically, we now emphasize the distinction between "social monitoring" (characteristic of unfamiliar dyads) and "active social coordination" (characteristic of familiar dyads).

(1) Topic Sentences: We revised the topic sentences of all Results paragraphs to immediately highlight the findings regarding male interest and the influence of familiarity on reciprocation.

(2) Conceptual Framework: We added a conceptual distinction in the Discussion, explaining that while unfamiliar marmosets maintain high social attention through "peripheral monitoring" and proximity-dependent joint gaze, familiar pairs exhibit sophisticated, distance-independent coordination and gaze reciprocation.

(3) Clarification of Male Interest: We explicitly stated that while male interest in females is high regardless of familiarity, it manifests as persistent monitoring in unfamiliar pairs versus a more aware, reciprocal state in familiar pairs.

Minor comments:

(1) Methods:

a) Lines 522-539: The 200 continuous frames used for validation of the model containing two marmosets are sufficient to test how well it generalizes to other animals outside the training set? The RMSE reported, does it vary for animals inside vs outside the training set? To what extent does the RMSE, in image pixels, translate into accuracy in estimating the gaze direction, for example, as assessed by estimating error when marmosets look at images or video on the monitor?

To address the reviewer’s concern regarding generalization and the translation of pixel RMSE to angular accuracy, we emphasize that the six facial features selected are prominent, high-contrast features across the species. Consequently, we observed that the RMSE remained consistent for marmosets both inside and outside the training set. To quantify how pixel-level tracking error translates into gaze estimation accuracy, we performed a sensitivity analysis. We simulated landmark (i.e., feature) jitter by sampling perturbations from circular distributions based on our empirical data (2.4 pixels for eyes; 2.1 pixels for the central blaze). Our results, illustrated in uthpr response image 1, show that 90% of the resulting head gaze deviations fall within 10°, which is consistent with the angular threshold used for our gaze cone model. This confirms that the reported RMSE provides sufficient precision for reliable gaze estimation.

Probability distribution of gaze angular deviation under circular perturbation.

The histogram (blue) represents the change in reconstructed gaze angle (degrees) following stochastic perturbation of facial features. To simulate real-world variance, noise was sampled from circular distributions with radii of 2.4 pixels (eyes) and 2.1 pixels (central blaze). The red curve represents an exponential fit to the empirical data (y=aebx, a=0.9591, b=0.1813. Approximately 90% of the reconstructed gaze deviations remain below 10°, indicating the model’s localised stability under pixel level coordinate jitter.

b) Line 542-43: Is there any difference between a rigid model fit to the six facial points, versus using the plane defined by the two eyes and central blaze in terms of direction accuracy (in the ground truth validation)? How does the "semi-rigid" set of six points (mentioned also in lines 201-203) constrain the fit of the three points (two eyes and central blaze) that define the normal plan for the gaze cone?

We thank the reviewer for the opportunity to clarify our geometric model. The plane used to define the gaze cone's origin was indeed determined by the two eyes and the central blaze. However, a plane defined by only three points was insufficient to determine a unique gaze direction, as the normal vector was ambiguous (it could point forward through the face or backward through the head).

To resolve this, we utilized the relative positions of the two ear tufts. Because the tufts are anatomically situated behind the eyes and blaze, these additional points provide the necessary spatial context to orient the gaze vector correctly. In our validation, we found that the mouth does not alter the angular accuracy compared to a 3-point fit, supporting that the facial features are correctly identified.

We use the term 'semi-rigid' to describe the six-point constellation because their relative spatial configurations remain stable across individuals and expressions, imposing a biological constraint on the model. This prevents unphysical warping of the face frame during 3D reconstruction and ensures the gaze cone remains anchored to the animal's true midline.

(2) Results:

a) Lines 203-205: What is the distinction between gaze orientation (defined by facial plane, 3D vector) and gaze direction (defined by ear tufts) ... is gaze direction in the 2D x-y plane? Why are two measures needed or different? It does not appear gaze orientation is used further in the manuscript and perhaps could be omitted.

We appreciate the reviewer’s comment regarding the terminology. We have replaced all instances of ‘gaze orientation’ with ‘gaze direction’ to ensure consistency throughout the manuscript.

To clarify, both terms referred to the same 3D unit vector. The ear tufts were not used to define a separate 2D measure; rather, they served as posterior anatomical anchors to resolve the 3D polarity of the normal vector (ensuring the vector points 'forward' from the face rather than 'backward'). Gaze direction was calculated in 3D space and was not restricted to a 2D x-y plane. We have clarified this in the revised Methods section (Lines 203–205) to avoid further ambiguity.

b) Line 215-216: why is head-gaze velocity put in normalized units instead of degrees visual angle per second? How was the normalization performed (lines 549-557)? It would be simpler to see velocity as an angular speed (degrees angle per second) rather than a change in norms.

We thank the reviewer for this suggestion. We agree that the expression is misleading.

(1) We have replaced "face norm" with "face normal vector" (N) throughout the manuscript to clarify that we are referring to the 3D unit vector perpendicular to the facial plane.

(2) Lines 224-225 and the corresponding Methods section (Lines 599-609) have been updated to reflect this change in units and terminology.

We chose to use the change in the face normal vector in normalized units for our primary calculations because it allows for efficient spatiotemporal smoothing and is computationally robust at the very low thresholds required for our stability analysis. However, to address the reviewer's concern regarding interpretability, we have verified that our threshold of 0.05 normalized units corresponds to an angular velocity of 2.87 degrees/frame duration [33ms]. Since we are operating at very small angular changes, the Euclidean distance between unit vectors is a near-linear proxy for the angular displacement in radians.

c) Lines 215-216: How do raw gaze traces appear over time ... are there gaze saccades and then stable fixations, or does it vary continuously? A plot of the gaze trace might be useful besides just showing velocity with a threshold, to evaluate to what extent stable fixation vs shifts are distinct.

Time course of gaze, angular velocity and stability, thresholding.

The plot illustrates the temporal dynamics of the face normal vector velocity used to define stable gaze states. The blue trace represents the raw gaze velocity calculated in normalised units. The red dashed line demotes the empirical cut off threshold of 0.05 units per frame.

To clarify the temporal dynamics of marmoset head movements, we have provided a representative time course of head gaze velocity as shown in Author response image 2. The data clearly show a "saccade-and-fixate" pattern: large, distinct spikes in velocity (representing rapid head redirections) are separated by periods of relative stability.

While minor high-frequency fluctuations in the raw trace (blue) may be attributed to facial feature detection noise, they remain significantly below our stability threshold (red dashed line). By applying this threshold, we successfully isolated biologically relevant "stable fixations" from "head saccades," ensuring that our subsequent social gaze analysis is based on periods of intentional head gaze direction.

d) Lines 237-286: The writing in this section does not emphasize the main results. There seem to be three takeaway points that could be emphasized better in the topic sentences of each of the paragraphs: i) Marmosets tended to spend most of their time on either end of the elongated box, not in the middle, ii) Males spent more time near the front of the box near the other animal than females, iii) Familiar pairs spent more time closer to each other.

To address this comment, we have reorganized this section to lead with the three key behavioral findings:

(1) We now state clearly in the topic sentence that marmosets preferred the ends of the arena over the middle.

(2) We have highlighted the finding that males spend significantly more time near the inner edge (closer to the partner) than females, irrespective of familiarity.

(3) We emphasized that familiar pairs maintain closer and more dynamic social distances over time, whereas unfamiliar pairs tend to move further apart as a session progresses.

e) Line 303: It would be useful to see time traces of head velocity of each member of the pair and categorization over time of the gaze event types. A stable epoch must be brief on the order of 100-200ms. It is unclear how distinct the stable fixation epochs are from the moments when the gaze is shifting. Also, the state transition analysis treats each stable epoch like one event, and then following a gaze movement by either of the pair, the state is defined again, is that correct?

We defined stable epochs as continuous periods where the face normal vector velocity remained below 0.05 normalized units for both animals. This ensures that a "gaze state" is only categorized when both marmosets have relatively fixed head orientations. As shown in the provided time traces in Author response image 2), the velocity profile is characterized by sharp peaks (head saccades) and clearly defined troughs (fixations). Further, we generated a probability histogram of stable head-gaze epoch durations (Author response image 3). The median duration of these stable epochs is 200ms, which aligns with biological expectations for fixation durations in primates and confirms that these states are distinct from the high-velocity shifts.

The reviewer’s interpretation is correct. Our Markov chain model treats each stable epoch as a single event. A transition occurs when at least one animal moves (exceeding the velocity threshold), resulting in a new stable epoch where the relative gaze state is re-evaluated. This approach allows us to model the sequence of social interactions as a series of discrete behavioral decisions.

Temporal characteristics of stable gaze, head gaze, epochs.

The histogram illustrates the probability distribution of the duration (ms) of stablegaze behaviour epochs. A minimum duration threshold of 100 ms was applied to exclude transient, non-purposeful head gazes.

f) Lines 316-326: Some general summarizing statements to lead this paragraph would be useful. It seems that familiar pairs are more likely to participate in joint gaze, especially when close to each other, and perhaps, that males tended to gaze at females more than the reverse. Is there any notion that males were following the gaze of females?

We thank the reviewer for these suggestions. We have revised the topic sentences of this section to lead with a summary of the social takeaways, specifically highlighting the higher level of male interest and the shift toward reciprocal coordination in familiar pairs.

The reviewer correctly identified an important dynamic. Our transition analysis (Fig. 4D) confirms that males in both familiar and unfamiliar dyads frequently follow the female's gaze. This is evidenced by a robust transition probability (~17%) from "Male-to-Female Partner Gaze" (blue node) to "Joint Gaze" (green node). We found that this gaze-following behavior was a general feature of the dyads and did not differ significantly by familiarity, which is why it was not previously emphasized. However, we have now added a statement to the Results (Lines 358-365) to explicitly describe this male-led gaze-following behavior.

g) Lines 328-337: Can these findings in this paragraph be summarized more generally? It seems males view unfamiliar females longer, whereas for familiar females they are more likely to reciprocate viewing if being viewed by them and then to join in joint gaze with them. Would that event, viewing a female and then a transition to joint gaze, not be categorized as a gaze-following event?

We have now summarized the paragraph to emphasize the transition from vigilant monitoring in unfamiliar pairs to reciprocal awareness in familiar pairs.

Regarding "longer" viewing: We have clarified the text to specify that males' interest in unfamiliar females is persistent and robust rather than simply "longer" in a single duration. The high recurrence probability signifies that males consistently re-orient their gaze back to the unfamiliar female even if the interaction is briefly interrupted by movement.

Regarding gaze following and joint gaze: The reviewer asks if the transition from viewing a female to joint gaze constitutes gaze following. We agree that a transition from "male-to-female gaze" to "joint gaze" is indeed a gaze-following event (as noted in our previous response regarding Fig. 4D). However, the specific transition discussed in this paragraph (female-to-male gaze to male-to-female gaze) is different: it describes a "reciprocal" event where the male responded to being looked at by looking back at the female, while the female simultaneously shifted her gaze away. Since the two gaze cones did not intersect on an external object or on each other's faces simultaneously at the end of this transition, it was not categorized as joint gaze or gaze following.

h) Lines 339-351: It is not clear why gazing at the region surrounding a female's face (as opposed to the face itself) reflects "gaze monitoring tied to increased social attention (Dal Monte et l, 2022). This hypothesis could be expanded to make the prediction clear in this paragraph.

We thank the reviewer for identifying the need to clarify the hypothesis regarding the region surrounding the face. We have expanded this paragraph to explain why gazing at the peripheral facial region reflects social monitoring.

In many primate species, direct and sustained eye contact can be often interpreted as a threat or a challenge, particularly between unfamiliar individuals. Peripheral monitoring (looking at the area immediately surrounding the face) can strategically allow an animal to stay highly attentive to the partner's head orientation, gaze direction, and facial expressions—all critical for anticipating future actions—while minimizing the risk of social conflict. By demonstrating that unfamiliar marmosets utilize this peripheral strategy significantly more than familiar ones, we provide evidence that social attention in novel dyads is characterized by a social monitoring strategy that balances the need for information with social caution.

i) Lines 354-373: This section seems to suggest again that in a familiar male/female pair, the male is more likely to follow the female gaze and establish a joint gaze, and this occurs less with the unfamiliar pair only when closer in distance. Some summary sentences to begin the paragraph could help frame what to expect from the results.

We have added summarizing topic sentences to this section to clarify the relationship between familiarity and the spatial distribution of joint gaze.

(3) Discussion:

Lines 380-463: This section reads more clearly than most of the results, where it is often hard to connect the data plots to their significance for behavior. Overall, I believe the manuscript could be improved by setting up a hypothesis before presenting results in the paragraphs demonstrating the data. Some of the main findings appear in text from lines 413-419 (somewhat hidden even in discussion).

We sincerely appreciate the reviewer’s positive feedback on the clarity of the latter sections of our Discussion. We have taken the suggestion to heart and have performed a comprehensive restructuring of the Results and Discussion sections.

(1) We have moved the key takeaways, specifically the distinction between vigilant monitoring in unfamiliar pairs and reciprocal coordination in familiar pairs, from the end of the Discussion to the topic sentences of the relevant Results paragraphs.

(2) We established a unified framework throughout the manuscript that connects pixel-level tracking stability to the biological "saccade-and-fixate" movement pattern, and ultimately to the social dimensions of sex and familiarity.

(4) A couple of additional questions to address in the discussion:

a) Can you speculate why in this behavioral context the marmosets do not engage in reciprocal gaze where both are simultaneously looking at each other (lines 297-301)? How low is the incidence of this event, numerically, in comparison to the other events (1 in 1000 events, etc)?

We appreciate the reviewer’s interest in the lack of reciprocal gaze (mutual eye contact).

Numerically, reciprocal gaze events occurred with a frequency of approximately 1 in 500 social gaze events (comprising less than 0.2% of our social dataset). Given this extreme scarcity, we felt that any statistical comparisons across sex or familiarity would be underpowered and potentially misleading, leading to our decision to focus on partner and joint gaze states.

We speculate that the rarity of reciprocal gaze is primarily due to our task-free experimental setup. Unlike directed cooperation tasks where animals must look at each other to coordinate actions for a reward (e.g., Miss & Burkart, 2018), our study focused on task-free interactions. In a free-moving context without a common goal, marmosets may prioritize monitoring the environment or the partner’s actions (joint or partner gaze) over direct, sustained mutual eye contact, which can sometimes be perceived as a confrontational or high-arousal signal in primate social hierarchies.

b) Does a transition from a marmoset viewing their partner, to a joint gaze, count as a gaze-following event? It appears the authors are reluctant to use that terminology. What are the potential concerns in that terminology? Is there a concern that both animals orient to the same object that is salient to them without it being due to their gaze?

A transition from a partner-directed gaze to a joint gaze is indeed a gaze-following event. We distinguish these events from a transition between partner-directed gazes (e.g., male-to-female to female-to-male). In these "reciprocation" cases, once the second animal looked at the first, the first animal shifted their gaze away. Because the two gaze cones did not intersect on a common object at the end of the transition, I classified such events as a social exchange of attention rather than a coordinated gaze-following event.

Reviewer #2 (Recommendations for the authors):

I do have a few questions/points for clarification:

(1) While your approach appears to be able to track head orientation when the face is occluded or turned away from the primary cameras, how was the accuracy of this validated? Since you have multiple cameras, it should be possible to make the estimate using the occluded cameras and then validate using the non-occluded ones.

We appreciate the reviewer's comment regarding the validation of our tracking during partial occlusions.

We wish to clarify that our system does not utilize "primary" vs "auxiliary" cameras. Rather, any two or more cameras that capture facial features with high confidence are used to triangulate the points into 3D space. Thus, the "primary" cameras are dynamically determined frame-by-frame based on the animal's orientation.

To validate the accuracy of our 3D reconstruction during occlusions, we utilized a "projection-validation" approach. As demonstrated in Figure 2B (left panel), when the face is turned away from a specific camera, leaving only the back of the head visible, we used the facial features triangulated from the other non-occluded cameras and projected them onto the image plane of the occluded camera. The fact that these projected points aligned precisely with the expected (but hidden) anatomical landmarks confirms the global accuracy of our 3D model.

We previously benchmarked this approach using a three-camera system where we triangulated coordinates via two cameras and successfully projected them onto the third camera's image plane with high accuracy. This ensures that even when a camera is "blind" to the face, the 3D position estimated by the rest of the array remains robust.

(2) Marmosets, like other non-human primates, also look at other body postures for their social communication, though admittedly marmosets are far more likely to look others in the face than larger primates. The tail-raised genital displays come to mind. While the paper primarily focuses on shared vs deviant gaze, and I believe tracks not only the angle of viewing towards the target but also the distance from the face (please clarify if I am wrong), it would also be useful to know how often marmosets are looking at each other beyond just the face. This is particularly interesting if the gaze towards the partner varies depending on whether that partner was generally oriented towards the gazer, or not. For the joint gaze, were there conditions in which the two were looking at the same target, but had body postures that were not oriented toward one another (i.e. looking at a distant target beyond one of the animals, like looking over someone else's shoulder)?

We thank the reviewer for highlighting the importance of body postures and non-facial social signals (e.g., genital displays) in marmoset communication.

At the inception of this project, we explored tracking multiple body parts. However, due to the marmoset's dense fur and the lack of distinct skeletal markers under naturalistic lighting, human annotators and early automated tools struggled to achieve the precision required for high-resolution 3D kinematics. While recent advances in whole-body tracking now make these questions approachable, we chose to focus on the face normal vector because it provided the most robust and high-confidence signal for social orientation in our current dataset.

Regarding the "looking over the shoulder" scenario, we utilized a hierarchical classification system to prevent wrong categorization. Intersection with the partner’s face always took priority. If one animal’s gaze cone contained the other’s face, the state was classified as "Partner Gaze", even if the two gaze cones happened to intersect at a distant point in space. This ensures that "Joint Gaze" specifically captures instances where both animals ignore one another’s face regions to focus on a shared external target.

We agree that the relationship between body posture and head gaze is a fascinating area for future research. In our current setup, while "Joint Gaze" requires the head-gaze cones to intersect, the animals' bodies could indeed be oriented in different directions (e.g., looking at a distant target behind the partner). We have added a note to the Discussion acknowledging that incorporating whole-body gestures would further deepen the understanding of marmoset social ethology.

(3) In the introduction, (line 70), you raise the question of ecological relevance, using rhesus in laboratory settings. This could use a little more expansion/explanation of the limitations of current/past approaches.

We thank the reviewer for the suggestion to expand upon the ecological limitations of traditional laboratory paradigms.

We have substantially revised the Introduction (Lines 70–82) to provide a more detailed critique of past approaches. Specifically, we now highlight how traditional head-fixed or screen-based paradigms decouple eye movements from natural head-body dynamics and lack the reciprocal, multi-agent complexity found in real-world social environments (e.g., Land, 2006; Shepherd, 2010). By contrasting these constraints with the spatially and socially embedded nature of marmoset interactions, we clarify why a more naturalistic, quantitative approach is necessary to understand the true dynamics of social gaze. These additions provide a stronger theoretical foundation for our move toward a free-moving experimental model.

Version 2

Abstract

Social communication relies on the ability to perceive and interpret the direction of others’ attention, and is commonly conveyed through head orientation and gaze direction in humans and nonhuman primates. However, traditional social gaze experiments in nonhuman primates require restraining head movements, significantly limiting their natural behavioral repertoire. Here, we developed a novel framework for accurately tracking facial features and three-dimensional head gaze directions of multiple freely moving common marmosets (Callithrix jacchus). By combining deep learning-based computer vision tools with triangulation algorithms, we were able to track the facial features of marmoset dyads within an arena. This method effectively generates dynamic 3D geometrical facial frames while overcoming common challenges like occlusion. To detect the head gaze direction, we constructed a virtual cone, oriented perpendicular to the facial frame. Using this pipeline, we quantified different types of interactive social gaze events, including partner-directed gaze and joint gaze to a shared spatial location. We observed clear effects of sex and familiarity on both interpersonal distance and gaze dynamics in marmoset dyads. Unfamiliar pairs exhibited more stereotyped patterns of arena occupancy, more sustained levels of social gaze across social distance, and increased social gaze monitoring. On the other hand, familiar pairs exhibited higher levels of joint gazes. Moreover, males displayed significantly elevated levels of gazes toward females’ faces and the surrounding regions, irrespective of familiarity. Our study reveals the importance of two key social factors in driving the gaze behaviors of a prosocial primate species and lays the groundwork for a rigorous quantification of primate behaviors in naturalistic settings.

Introduction

Primates, including humans, exhibit complex social structures and engage in rich interactions with members of their species, which are crucial for their survival and development. Among social stimuli, the face holds paramount importance with specialized neural systems (Deen et al., 2023; Hesse & Tsao, 2020) and is attentively prioritized by primates during much of their social interaction. Notably, the eyes garner the most attention among all facial features, playing a pivotal role in indicating the direction of others’ attention and also possibly their intention (Dal Monte et al., 2015; Emery, 2000; Itier et al., 2007). Indeed, understanding and interpreting the gaze of fellow individuals is a fundamental attribute of the theory of mind (ToM) (Martin & Santos, 2016; Saxe & Kanwisher, 2003). While current studies of social gaze using either realistic stimuli or pairs of rhesus macaques (Macaca mulatta) in a controlled laboratory setting (Dal Monte et al., 2016; Mosher et al., 2014; Ramezanpour & Thier 2020; Shepherd et al., 2006; Shepherd & Freiwald, 2018) provide valuable insights into social gaze behaviors, they are nevertheless limited in their ecological relevance. Many laboratory paradigms rely on head-restrained animals and simplified stimuli, decoupling eye movements from natural head–body dynamics and limiting spontaneous social behavior (Land, 2006; Einhauser et al., 2007; Foulsham et al., 2011). Social gaze is also typically examined in tightly controlled dyadic or screen-based paradigms, which lack the reciprocal, context-dependent, and multi-agent dynamics of natural social interactions (Shepherd, 2010; Birmingham et al., 2008). In contrast, gaze behavior in natural settings is shaped by locomotion, body posture, spatial relationships, and social structure—factors largely absent from laboratory studies (Emery, 2000; Klein et al., 2009). Consequently, the generalizability of findings from controlled experiments to freely moving, socially embedded interactions remains unclear, motivating the need for more naturalistic yet quantitative approaches.

To address these limitations, we turned to common marmosets (Callithrix jacchus), a highly prosocial primate species known for their social behavioral and cognitive similarities to humans (Miller et al., 2016). Marmosets are also a model system with increasing applications in computational ethology (Mitchell et al., 2014; Ngo et al., 2022). Like humans, they engage in cooperative breeding, a social system in which individuals care for offspring other than their own, usually at the expense of their reproduction (French, 1997; Solomon & French, 1997). Gaze directions, inferred from head orientation, hold crucial information about marmoset social interactions (Heiney & Blazquez, 2011; Spadacenta et al., 2022). The emergence of computational ethology (Anderson & Perona, 2014; Datta, Anderson, Branson, Perona, & Leifer, 2019) has propelled the development of a host of computer vision tools using deep neural networks (e.g., OpenPose by Cao et al., 2017, DeepLabCut by Mathis et al., 2018, DANNCE by Dunn et al., 2021; SLEAP by Pereira et al., 2022). Moreover, we have recently developed a scalable marmoset apparatus for automated pulling (MarmoAAP) for studying cooperation in marmoset dyads (Meisner et al., 2024). However, tracking head gaze direction in multiple freely moving marmosets poses a challenging problem, not yet solved by existing computer vision tools. This problem is further complicated by the fact that accurately tracking gaze directions in primates requires three-dimensional information.

Here, we propose a novel framework based on a modified DeepLabCut pipeline, capable of accurately detecting body parts of multiple marmosets in 2D space and triangulating them in 3D space. By reconstructing the face frame with six facial points in 3D space, we can infer the head gaze of each animal across time. Importantly, marmosets that are not head-restrained use rapid head movements for reorienting and visual exploration (Pandey et al., 2020), and therefore the head direction serves as an excellent proxy for gaze direction in unrestrained marmosets. With this framework in place, we investigated the gaze behaviors of male-female pairs of freely moving and interacting marmosets to quantify their social gaze dynamics. We investigated these gaze dynamics along the dimensions of sex and familiarity and found several key differences along both of these important social dimensions, including increased partner gaze in males that is modulated by familiarity, increased joint gaze among familiar pairs, and increased gaze monitoring by males. This fully automated tracking system can thus serve as a powerful tool for investigating primate group dynamics in naturalistic environments.

Results

Experimental setup and reconstruction of video images in 3D space

The experimental setup consisted of two arenas made with acrylic plates that allowed two marmosets to visually interact with each other while being physically separate (Fig. S1A). Each arena was 60.96 cm long, 30.48 cm wide, and 30.48 cm high. Five sides of the arena, except the bottom, were transparent allowing a clear view of the animal subjects under observation. The bottom side of the arena was perforated with 1-inch diameter holes arranged in a hexagonal pattern to aid the animal’s traction. The arenas were mounted on a rigid frame made of aluminum building blocks, with the smaller sides facing each other, and were separated by a distance of 30.48 cm. A movable opaque divider was placed between the arenas during intermittent breaks to prevent the animals from having visual access to each other (Methods). Two monitors were attached to the aluminum frame, one on each end, for displaying video or image stimuli to the animals. To capture the whole experimental setup, two sets of four GoPro 8 cameras were attached to the frame, where each set of cameras captured the view of one of the arenas.

After obtaining the intrinsic parameters of the cameras by calibration and the extrinsic parameters of the cameras by L-frame analysis (see Methods), we established a world coordinate system of each arena surrounded by the corresponding set of four cameras. Crucially, the two independent world coordinate systems of the two arenas were combined by measuring the distance between the two L-shaped frames and adding this offset to one of the world coordinate systems.

With the established world coordinate system, any point captured by two or more cameras could be triangulated into a common three-dimensional space. Thus, the experimental setup was reconstructed into three-dimensional space by manually labeling the vertices of the arenas and monitors in the image space captured by the cameras (Fig. S1B). The cameras on the monitor ends (marked as ‘ML1’, ‘ML2’, ‘MR1’, ‘MR2’ in Fig. S1B) recorded both animal subjects, whereas the cameras in the middle (marked as ‘OL1’, ‘OL2’, ‘OR1’, ‘OR2’ in Fig. S1B) recorded only one animal subject.

Automatic detection of facial features of two marmosets

Six facial features – the two tufts, the central blaze, two eyes, and the mouth (Fig. 1A) – were selected for automated tracking using a modified version of a deep neural network (DeepLabCut, Mathis et al., 2018). The raw video from each camera was fed into the network to compute the probability heatmaps of each facial feature. We modified the original method to detect features of two animals (Fig. 1B). After processing the raw video, two locations with the highest probability over a threshold (95%) were picked from each probability heatmap (Fig. 1B, feature detection). Since all the features from the same animal should be clustered in image space, a K-means clustering algorithm (Fig. 1B, initial clustering) was used on the candidate features with the constraint that one animal can only have one unique feature (Fig. 1B, refine clustering); for example, one animal cannot have two left eyes. After clustering, two clusters of features corresponding to the two animals were obtained. To detect outliers that were not valid features, we first calculated a distribution of within-cluster distances (Fig. 1B, remove outliers). Outliers were determined as those points that had nearest-neighbor distances which were two standard deviations above the average within-cluster distance and were excluded from subsequent analyses. Note that the above analyses were performed independently for each video frame.

Pipeline of detecting facial features of two marmosets.

(A) Six facial features of the marmoset (face frame) are color-coded: right tuft (red), central blaze (yellow), left tuft (green), right eye (purple), left eye (blue), and mouth (magenta). (B) Feature tracking pipeline (right) with the corresponding illustration for each step across two adjacent video frames (left). At the end of the pipeline, the facial features are clustered with the identities assigned consistently across frames. Facial points are color-cod­ed as in A. (C) Example frames of four steps in the pipeline shown in B. It can be seen clearly that the facial points are tracked and clustered accurately, and the identities are consistent across frames.

To establish temporal continuity across video frames and track animal identities, we first calculated the centroid of each cluster (Fig. 1B, calculate centroids). Under the heuristic that centroid trajectories corresponding to each individual animal are smoothly continuous in space and time (i.e., there are no sudden jumps or reversals in centroid location across frames at our sampling rate of 30Hz), we assigned identities to the centroids thus enabling us to track identities over time (Fig. 1B, establish identities). The facial features corresponding to a centroid inherited this identity (Fig. 1B, cluster with identities). We tested our automated pipeline against a manually labeled data set and found that features were detected with high precision (root mean squared error = 3.31 ± 0.1 pixels [mean +/- s.e.m.]; n = 2400 facial features; see Methods).

Our method thus allowed us to accurately detect and track the facial features of two marmosets in the arena (Fig. 1C), and in more general contexts such as in a home cage with occlusions (Supplementary Video 1).

Inferring head-gaze direction in 3D space

In our experimental setup, the cameras in the middle unambiguously recorded only one animal. The centroids of the facial features in the image space recorded by the middle cameras were triangulated into three-dimensional space and were constrained to be confined within the bounds of the arena. Any missing centroids were filled by interpolation from neighboring frames from both the past and future time points. The triangulated centroids acquired from cameras in the middle were then projected into the image space of cameras on the monitor ends. Since both animals were recorded by the cameras on the monitor ends, facial features from these camera views were detected with identities assigned (as described in the 2D pipeline). In the image space of cameras on the monitor ends, the cluster of facial feature points closer to the projected centroid and within the bounds of the arena were kept for later triangulation.

The detected facial features captured by all four cameras for one animal were subjected to triangulation. For each feature, results from all possible pairs of four cameras were triangulated. All the triangulation results were averaged to yield the final coordinates of the body part in three-dimensional space. Any missing features were filled by interpolation from neighboring frames including the previous and future time points. Testing against a manually labeled and triangulated data set yielded a high precision of 3D feature detection (root mean squared error = 3.7 ± 0.1 mm [mean +/- s.e.m.]; n = 2400 facial features; see Methods).

The six facial points constituted a semi-rigid geometric frame as the animal moves in 3D space (‘face frame’; Fig. 1), allowing us to infer the animal’s head-gaze direction as follows. The gaze direction was calculated as the normal to the facial plane (‘face norm vector’) defined by the two eyes and the central blaze. The position of the ear tufts, which were behind this facial plane, was used to determine gaze direction. Since marmoset saccade amplitudes are largely restricted to 10 degrees (median less than 4 degrees) (Mitchell et al., 2014), we modeled the head gaze as a virtual cone with a solid angle of 10 degrees (‘gaze cone’) emanating from the facial plane (Fig. 2A). Notably, with multiple camera views, the face frame can be reconstructed even when the face was invisible to one of the cameras, such that the reconstructed face frame in 3D can be projected back into the image space to validate the accuracy of the detection and reconstruction (Fig. 2B). We were thus able to obtain the animal’s continuous movement trajectory and the corresponding gaze direction over time (Fig. 2C, Supplementary Video 2). To characterize stable gaze epochs for subsequent analyses, we calculated the velocity of the face norm vector and applied a stringent threshold of 0.05 (normalized units) below which the marmoset’s head movement was considered stationary (Fig 2D).

3D Reconstruction of facial features and head gaze modeling.

(A) The face frames of two marmosets are reconstructed in 3D using the tracked facial points in Fig. 1A. A cone perpendicular to the face frame (gaze cone; 10-degree solid angle) is modeled as the head gaze. (B) Two example frames with the facial points projected from 3D space onto different camera views are shown. The left frame demonstrates that the facial points can be detected using information from other cameras, even if the face is invisible from that viewpoint. (C) Trajectory of the reconstructed face frame and the correspond­ing gaze cones across time. (D) The histogram of the face norm velocity. The red dotted line (0.05) is the threshold below which the marmoset head direction is considered to be stationary

We first examined our method’s ability to characterize gaze behaviors of freely moving marmosets by presenting either video or image stimuli to individual animals on a monitor screen (see Methods). We analyzed gaze behavior following the onset of video stimuli presented at different monitor locations (Fig. S2A). The analysis was restricted to video clips in which human annotators confirmed that the marmosets were looking at the monitor. Consistent with prior work in head-fixed marmosets (Mitchell et al., 2014), gaze–monitor intersection centers clustered around the corresponding stimulus locations after stimulus onset, supporting the spatial accuracy of the estimated gaze directions. Marmosets exhibited longer gaze duration to video stimuli compared to image stimuli (Fig. S2B; Mann Whitney U Test, p < 0.001). However, this difference was not caused by differences in gaze dispersion (Fig. S2C; Mann Whitney U Test, ns). By examining the frequency of gaze events, we found that marmosets gazed at the monitor more during the early period of video stimuli presentations compared to the late period (Fig. S2D; Mann Whitney U Test, p < 0.001), while there was no such difference for the image stimuli (Fig. S2D; Mann Whitney U Test, p = 0.4345). There was also a significant difference in gaze frequency between the early period of video presentations compared to the same period for image presentations (Fig. S2D; Mann Whitney U Test, p < 10-10). Taken together, our results support that dynamic visual stimuli elicit greater overt attention compared to static stimuli in marmosets, similar to macaques (Dal Monte et al., 2016; Furl et al., 2012) and humans (Chevallier et al., 2015).

Positional dynamics of marmoset dyads

With this automated system in place, we recorded the behavior of four pairs of familiar marmosets and four pairs of unfamiliar marmosets. We first observed that preferred spatial position was highly non-uniform, with animals preferring to occupy the ends of the elongated arena rather than the center. Data from each pair was recorded in one session consisting of ten five-minute free-viewing blocks interleaved with five-minute breaks. Each pair consisted of a male and a female animal. The familiar pairs were defined as cage mates, while each member of an unfamiliar pair was from a different home cage with no visual access to each other while in the colony. We first examined the movement trajectories of marmoset dyads and used the centroids of the face frames across time to represent the trajectories. For an example 5-minute segment (Fig. 3A), we observed that the marmosets preferred to stay at the two ends of their respective arenas. This was confirmed by a heatmap of projections of the trajectories on the plane parallel to the vertical long side of the arenas (‘XZ’ plane). Furthermore, there were two hotspots along the vertical axis in the heatmap of projections to the vertical short side plane (‘YZ’ plane), suggesting that the animals’ preferred body postures were either upright or crouched.

Positional dynamics of marmoset dyads.

(A) Movement trajectories of the face frame centroids for a marmoset pair (orange for female, cyan for male) in an example five-minute block. The heatmaps were calculated using the projections of the trajectories to XY, YZ, and XZ planes. (B) Marginal distributions of movement trajectories along the X and Z axes were calculat­ed for all marmosets and grouped by familiarity and sex (transparent colors for familiar pairs, opaque colors for unfamiliar pairs). Black bars indicate significant differences between pairs of distributions (Mann-Whitney U test, significance level at 5%). (C) Histograms of social distance along the X axis show trimodal distributions for both familiar pairs (gray) and unfamiliar pairs (black). The top inset shows a schematic of the arena config­uration with corresponding colors (orange for females and cyan for males). Social distance is defined as the Euclidean distance between the two marmosets. The fitted red curve for the unfamiliar pairs is a tri-Gaussian distribution, while the fitted red curve for the familiar pairs is a mixture of Gamma and Gaussian distributions, with the first peak as the Gamma distribution. The three regions were designated as ‘Near’, Intermediate’, and ‘Far’. The bottom inset illustrates the reason for this nomenclature. (D) Temporal evolution of social distance for unfamiliar and familiar pairs (which each 5-minute viewing block). The central dark line is the mean and the shaded area is the standard deviation. Black dots indicate significant differences (Mann-Whitney U test, significance level at 5%).

Second, we found a distinct sexual dimorphism in spatial positioning, where males consistently stayed closer to their partners than females. We examined the marginal distributions of the movement trajectories along the horizontal (‘X’) and vertical (‘Z’) axes across all sessions and grouped them along the dimensions of sex and familiarity (Fig. 3B). Along the X axis (Fig. 3B, left), the distributions were slightly bimodal, with the main peak in the region near to the inner edge of the arenas. Regardless of familiarity, male marmosets tended to stay closer to the inner edge compared to females, as shown by the significant differences in the distributions when X ranged from 0 to 150 mm. However, there were no significant differences between the same sex members of familiar and unfamiliar pairs. For the Z axis (Fig. 3B, right), the distributions were bimodal (Bimodality coefficients (BC) all exceeded the threshold at 5/9 with a 5% margin; BC(familiar male) = 0.5863; BC(familiar female) = 0.6278; BC(unfamiliar male) = 0.6235; BC(unfamiliar female) = 0.6570; Warren Sarle’s bimodality test), consistent with what we observed in the heatmaps indicating either upright or crouched postures. The positional distributions along the Z axis were not different based on sex or familiarity.

Third, our analysis of dyadic distance revealed that familiar pairs spend more time in close proximity and maintain more flexible social distances over time compared to unfamiliar pairs. To characterize the social distance dynamics of the freely moving dyads, we calculated the distance between the centroids of the pairs. We then examined the distributions of the social distance along the X-axis separately for familiar and unfamiliar pairs (Fig. 3C). The distributions were trimodal, and can be explained by the bimodal distribution of movement trajectories of individual marmosets along the X-axis. As mentioned above, marmosets tended to stay at the two ends of their arenas, and thus, combinations of preferred positions at the two ends for the dyads (see insets in Fig. 3C) resulted in the trimodal distribution. We termed these three peaks as ‘Near’, ‘Intermediate’, and ‘Far’. To quantify these distributions, we fitted the empirical data with mixture models using maximum likelihood estimation (see Methods). The social distance for unfamiliar pairs was best fitted by a mixture of three Gaussians while the distribution for familiar pairs was best fitted by a mixture of Gamma and Gaussian distributions. The first peak (‘Near’) of the familiar-pair distribution was best fitted by a Gamma distribution, implying a higher degree of dispersion when familiar marmosets were close to each other. Upon examining the temporal evolution of the social distance (within each 5-minute viewing block), we found that the social distance of unfamiliar pairs increased over time, whereas this distance fluctuated over time for familiar pairs (Fig. 3D), further indicating that the social distance dynamics of marmoset dyads depended on familiarity.

Social gaze dynamics of marmoset dyads

We next investigated the interactive aspects of gaze behaviors in freely moving marmoset dyads. We found that social interaction in marmosets is characterized by a distinct sexual dimorphism in gaze interest and a transition from “monitoring” in unfamiliar pairs to “active reciprocation” in familiar pairs. The gaze interaction between two animals could be simplified as the relative positions of two gaze cones in three-dimensional space (see Methods; Fig. 4A; Supplementary Video 3). If the gaze cone of one animal intersected with the facial plane of the second animal (but not vice versa), we termed it ‘partner gaze’. If the gaze cones of both animals intersected with that of the other’s facial place, we termed it ‘reciprocal gaze’. In our dataset, the instances of reciprocal gaze were very low and were thus excluded from further analysis. If the two cones intersected anywhere outside the facial planes, we termed it ‘joint gaze’. All other cases were regarded as ‘no interaction’ between the two animals.

Live interactive gaze analysis of unfamiliar and familiar marmoset dyads.

(A) Gaze type categorized based on the relative positions of the gaze cones. Joint gaze is defined as two marmosets looking at the same location. A partner gaze is defined as one animal looking at the other animal’s face (but not vice versa). No interaction occurs when the two gaze cones do not intersect. (B) Histograms of gaze count as a function of inter-animal distance, shown separately for familiar and unfamiliar pairs. (C) Same data as in B shown as pie charts of percentages in social gaze states. (D) Gaze state transition diagrams for familiar and unfamiliar pairs. The nodes are the gaze states and the edges connecting the nodes represent the transition between states. Edge colors indicate transition probabilities. (E) Partner gaze self-transition probabilities for familiar and unfamiliar pairs (χ2 test). (F) Delta transition matrix between the unfamiliar pair and familiar pair state transition diagrams. Transitions that are significantly different across familiarity are marked by asterisks (χ2 test, male to male, p < 0.05; female to male, p < 0.05; joint to joint, p < 0.0001). (G) Left, The schematic illustrates how gazing toward the surrounding region of a partner’s face area was measured. Right, Counts of gaze towards the surrounding region of the partner’s face by familiarity and sex. (Mann-Whitney U test, “* means p < 0.001; **‘* means p < 0.0001)

Our analysis of gaze states reveals that male marmosets maintain a high baseline of interest in females, though this relationship depends on familiarity and spatial proximity. We identified stable gaze epochs by thresholding the face norm vector velocity (Fig. 2D). Stable epochs were categorized into gaze states based on the gaze event types described above. We first analyzed the fraction of gaze states in the three position ranges identified from the social distance analysis (Fig. 4B,C). We found that male marmosets gazed more toward their partner females’ faces regardless of familiarity (p < 0.01, χ2 test). However, while male interest in familiar females decreased as social distance increased, interest in unfamiliar females remained constant across different social distances, suggesting a persistent monitoring of novel partners (Fig. 4C). Moreover, females in unfamiliar pairs exhibited significantly higher partner gazes (female→male) compared to those in familiar pairs (p < 0.01, χ2 test; Fig. 4C). The total counts of social gaze states (joint gaze and partner gaze) were higher for familiar pairs when they were near, but these decreased more dramatically with increasing distance (Fig. 4B).

Our state-transition analysis revealed that gaze dynamics are governed by a combination of persistent male attention and a shared tendency for males to follow the female’s gaze into joint attention. To investigate the dynamics of these gaze states, we computed state transition probabilities among distinct gaze event types for familiar and unfamiliar dyads. Markov chain modeling of state transitions further highlighted that familiar dyads engage in more complex, reciprocal social exchanges compared to the one-sided monitoring seen in unfamiliar dyads (Fig. 4D). We first focused on the recurrent (self-transition) edges for the partner gaze states. Recurrent edges indicate a transition back to the same stable gaze state after a break likely due to physical movement, and reflect the robustness of the state despite movement. In line with our previous results (Fig. 4B,C), males exhibit significantly higher recurrent partner gazes compared to females, irrespective of familiarity (χ2 test, unfamiliar male vs unfamiliar female, p < 10-10; familiar male vs familiar female, p < 0.01; Fig. 4E). Furthermore, we observed a consistent pattern of gaze- following across both groups; approximately 17% of male-to-female partner gazes transitioned directly into joint gaze (Fig. 4D), suggesting that males frequently utilize the female’s gaze orientation to guide their own environmental exploration.

Our comparison of state transition probabilities reveals that familiarity transforms male gaze behavior from a persistent monitoring of novel partners to a more socially responsive, reciprocal interaction (Fig. 4F). First, recurrent male partner gaze (male→female) was significantly enhanced in unfamiliar pairs (p < 0.05, χ2 test), suggesting a heightened interest in unfamiliar females. Second, there was a higher probability of transition from a female partner gaze to a male partner gaze in familiar pairs compared to unfamiliar pairs, suggesting that familiar males have a greater awareness of and tendency to reciprocate their partners’ gaze (p < 0.05, χ2 test). Third, there was a higher probability of recurrent joint gazes in familiar pairs compared to unfamiliar pairs, suggesting that familiar pairs explore common objects more than unfamiliar pairs (p < 10-4, χ2 test).

Monitoring others to anticipate their future actions is critical for successful social interactions (Hari et al., 2015). In particular, successful interactive gaze exchanges require constant monitoring of others’ gaze. In addition to direct eye contact, we found that unfamiliar marmosets engage in increased peripheral monitoring of their partners. We analyzed the gaze distribution in the surrounding region of a partner’s face to estimate gaze monitoring tied to increased social attention (Dal Monte et al., 2022). We quantified this by the distance between the centroid of the partner’s face-frame and the point of intersection of the gaze cone with the partner’s facial plane (Fig. 4G, left). Unfamiliar marmosets (both males and females) showed significantly higher (Mann-Whitney U test, unfamiliar male vs familiar male, p < 0.0001; unfamiliar female vs familiar female, p < 0.001) incidences of gaze toward the surrounding region of the partner’s face (Fig 4G, right; compare darker lines with the lighter lines). Further, males exhibited markedly higher (Mann-Whitney U test, p < 0.0001) incidences of gaze toward the partner females’ face (Fig 4G, right; compare cyan lines with the orange lines).

Finally, our analysis of joint gaze distribution underscores that familiar pairs establish shared attention more flexibly across their social distance, whereas shared attention in unfamiliar pairs are constrained by spatial proximity. Joint gaze is crucial in primates as it underpins social communication and coordination, serving as a foundation for more complex behaviors like cooperation and shared attention (Emery, 2000; Tomasello et al., 2005). We analyzed joint gaze behaviors between marmoset dyads, first projecting the locations of joint gazes onto different 2D planes around the arena (Fig. 5A). Significant asymmetry was observed in the distribution of joint gazes on the ‘XY’ and ‘XZ’ planes, with the majority of joint gazes occurring within the female’s arena, regardless of familiarity. This is consistent with our earlier results showing more partner-directed gazes from males and increased attention to the region surrounding the female partner (Fig. 4B,C,G). Notably, joint gazes between familiar pairs were more centrally distributed compared to those between unfamiliar pairs. When examining the ‘XZ’ and ‘YZ’ planes, we found that most joint gazes were concentrated along the lower Z-axis, indicating that marmoset pairs tend to establish shared attention closer to ground level. Given the importance of social distance in shaping marmoset behaviors, we further investigated the distribution of joint gazes as a function of social distance. Striking differences were found between familiar and unfamiliar pairs (Fig. 5B). The distribution for familiar pairs was best fitted by a tri- Gaussian model, consistent with the distribution of social distances observed in Fig. 3C, suggesting that familiar pairs can establish joint gazes regardless of distance. In contrast, the distribution for unfamiliar pairs was best described by a lognormal model, indicating that unfamiliar pairs tend to establish joint gazes only when in close proximity.

Joint gaze analysis between familiar and unfamiliar dyads.

(A) Heatmaps of joint gaze locations projected onto 2D planes (From left to right: ‘XY’, ‘XZ’, ‘YZ’). The icon on the left show a schematic of a joint gaze along with the arena configuration with corresponding colors (orange for females and cyan for males). The colored rectangles superimposed on the heatmaps show the projections of the arenas onto those 2D planes. The white dotted line indicates the midline between the two arenas. (B) The probability of joint gaze at varying social distances shows distinct distributions for unfamiliar pairs (black) and familiar pairs (gray). The fitted red curve for the unfamiliar pairs is a lognormal distribution, while the fitted red curve for the familiar pairs is a tri-Gaussian distribution. Thin fitted red curves in both panels are the data from individual pairs. The inset shows a schematic of a joint gaze and the social distance metric. Social distance is defined as the Euclidean distance between the two marmosets.

Overall, we found that both the social dimensions we examined – familiarity and sex – are significant determinants of natural gaze dynamics among marmosets.

Discussion

In this study, we first presented a novel framework for the automated, markerless, and identity-preserving tracking of 3D facial features of multiple marmosets. By building on top of the deep-learning framework provided by DeepLabCut, we used a constellation of cameras to overcome “blindspots” due to occlusion and imposed spatiotemporal smoothness constraints on the detected features to establish and preserve identities across time. The tracked facial features from each animal form a semi-rigid face frame as the animal moves freely in 3D space, thereby allowing us to infer the animal’s gaze direction at each moment in time. It is important to reiterate that unrestrained marmosets use rapid saccadic head-movements for reorienting (Pandey et al., 2020) and have a limited amplitude range of saccadic eye-movements (Mitchell et al., 2014). Thus their head direction serves as excellent proxy for gaze direction in unrestrained conditions.

Current methods for tracking and estimating animal body postures, such as DeepLabCut (Mathis et al., 2018) and SLEAP (Pereira et al., 2022), offer substantial advantages in precision and flexibility, particularly through the use of deep learning for markerless tracking across diverse species. These methods excel in laboratory settings, where they can achieve high accuracy without invasive markers and are adaptable for various animals and tasks. However, many of these tools struggle with occlusion, especially in crowded environments or natural habitats, limiting their effectiveness in complex social interactions or open-field studies (Sturman et al., 2020). Additionally, when tracking marmosets that exhibit significant vertical movement, requiring 3D tracking, the performance of these methods is suboptimal. While 3D tracking systems like DANNCE (Dunn et al., 2021) offer enhanced spatial accuracy, these methods are limited to single- animal tracking, which is insufficient for studies in social neuroscience.

Primates are highly visual species whose physical explorations of their environment are not confined to two-dimensional surfaces. Gaze is a critical component of primate social behavior and conveys important social signals such as interest, attention, and emotion (Emery, 2000). Assessment of gaze is therefore important to understand non-verbal communication and interpersonal dynamics. Our 3D gaze tracking approach was able to capture both the positional and gaze dynamics of freely moving marmoset dyads in a naturalistic context. We observed clear effects of sex and familiarity on both interpersonal and gaze dynamics. Unfamiliar pairs exhibited more stereotyped patterns of arena occupancy, more sustained levels of social gaze across distance, and increased gaze monitoring, suggesting elevated levels of social attention and the need to constantly track the movements of a novel conspecific compared to familiar pairs. On the other hand, familiar pairs exhibited more recurrent joint gazes in the shared environment compared to unfamiliar pairs. Critically, familiar males also showed a higher tendency to reciprocate their partner’s gaze, suggesting a greater awareness of their partner’s social gaze state and a transition from mere monitoring to active social engagement.

Supported by the natural ethology of marmosets (Yamamoto et al., 2014; Solomon & French, 1997), we found dramatic sex differences in gaze behaviors, with males exhibiting significantly elevated levels of gaze toward females’ faces and the surrounding regions irrespective of familiarity. It is important to note that dominance in marmosets is not strictly determined by gender, as it can vary based on individual personalities and intra-group social dynamics, although breeding females typically dominate social activity within a group (Digby, 1995; Mustoe et al., 2023). While we have not explicitly controlled for dominance in this study, whether part of the observed differences can be attributed to dominance effects needs further exploration.

The distinction between social monitoring and social coordination is further reflected in the spatial flexibility of joint attention. Gaze following plays a crucial role in social communication for humans and nonhuman primates, allowing for joint attention (Emery et al., 1997; Brooks & Meltzoff, 2005; Burkart & Heschl, 2006; Shepherd, 2010). Previous research demonstrated that head-restrained marmosets exhibited preferential gazing toward marmoset face stimuli observed by a conspecific in a quasi-reflexive manner during a free-choice task (Spadacenta et al., 2019). Here we found that males consistently followed the gaze of females into states of joint attention. This ‘gaze following’ occurred with similar frequency in both familiar and unfamiliar pairs, suggesting that monitoring a partner’s gaze to coordinate shared attention is a fundamental component of marmoset social ethology, regardless of the strength of the social bond. Interestingly, we did not find any differences in gaze-following behaviors (transition from partner gaze to joint gaze) along the social dimensions we tested here. Future investigation of such behaviors by manipulating social variables such as dominance or kinship could provide a comprehensive understanding of gaze following and joint attention in naturalistic behavioral contexts. The scarcity of reciprocal gazes in our study may be attributed to the task-free experimental setup employed. Indeed, in other joint action tasks requiring cooperation for rewards, marmosets actively engage in reciprocal gaze behaviors (Miss & Burkart, 2018).

While we focused on the tracking of facial features in this study, our automated system has the potential to extend to 3D whole-body tracking, encompassing limbs, tail, and the main body features of marmosets. In our system, multiple cameras surrounding the arena ensure that each body part of interest can be tracked through at least two cameras, enabling triangulation in 3D space. Our current system uses a pre-trained ResNet model (He et al., 2015) to track body parts of interest. However, considering the challenges posed by whole-body tracking, such as interference from marmosets’ fur that complicates feature detection, the adoption of cutting-edge transformer networks like the vision transformer model (Dosovitskiy et al., 2020) might significantly improve detection performance. Such an advancement in tracking and reconstructing the entire marmoset body frame would enable the analysis of such data using unsupervised learning techniques (Berman et al., 2014; Calhoun et al., 2019) and thereby provide a deeper understanding of primate social behavior.

In summary, our study lays the groundwork for a rigorous quantification of primate behaviors in naturalistic settings. Not only does this allow us to gain deeper insights beyond what is possible from field notes and observational studies, but it is also a key first step to go beyond current reductionist paradigms and understanding the neural dynamics underlying natural behaviors (Miller et al., 2022).